---
title: "Frío invernal en la Región de Murcia — mapas del test-run (Búlida / Búlida Precoz)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    fig_width: 11
    fig_height: 6
---

```{r setup, include=FALSE}
# Report that turns the PNACC daily temperature slice into winter-chill maps for the Murcia
# test-run. Chill is computed by compute_chill_table() (06_chill_murcia.R); everything below is
# the geographic presentation: CORINE cropland backdrop, Region de Murcia boundary (mapSpain),
# Safe Winter Chill per station, and the case-study classification against each cultivar's CR.
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, dpi = 130)

suppressWarnings(suppressMessages({
  library(data.table); library(sf); library(terra); library(mapSpain)
  library(ggplot2); library(viridis); library(patchwork)
}))

# --- paths -------------------------------------------------------------------------------
# Resolved by 00_paths.R. This file is knitted rather than run with Rscript, so `--file=` is not
# available and the location has to come from knitr::current_input(), which 00_paths.R handles.
.f <- grep("^--file=", commandArgs(FALSE), value = TRUE)
.d <- if (length(.f)) dirname(normalizePath(sub("^--file=", "", .f[1]))) else {
  ci <- try(knitr::current_input(dir = TRUE), silent = TRUE)
  if (!inherits(ci, "try-error") && !is.null(ci)) dirname(normalizePath(ci)) else getwd()
}
source(file.path(.d, "00_paths.R"))

TAB       <- plinius_data("tables", "murcia")
SLICE_DIR <- file.path(TAB, "chill_slice")
CHILL_CSV <- file.path(TAB, "chill_murcia.csv")
STATIONS  <- file.path(TAB, "stations.csv")
STRICT    <- file.path(TAB, "stations_cultivable_strict.csv")
BROAD     <- file.path(TAB, "stations_cultivable_broad.csv")
CLC       <- plinius_clc()
dir.create(FIG_DIR, recursive = TRUE, showWarnings = FALSE)   # FIG_DIR comes from 00_paths.R

CR_PRECOZ <- 33.7   # chill requirement of 'Bulida Precoz' (chill portions)
CR_BULIDA <- 47.5   # chill requirement of 'Bulida'

# equal-area CRS for the whole map (CORINE is already EPSG:3035)
EPSG <- 3035
```

```{r compute-chill}
# Load the chill table computed by 06_chill_murcia.R (DM_JOSE, Egea's 1987 dynamic model);
# otherwise compute it here. Not cached: chill_murcia.csv is the single source of truth, so a
# re-run of the compute script is picked up on the next knit.
if (file.exists(CHILL_CSV)) {
  chill <- fread(CHILL_CSV)
} else {
  source(file.path(SCRIPTS_DIR, "06_chill_murcia.R"))
  chill <- compute_chill_table(SLICE_DIR)
  fwrite(chill, CHILL_CSV)
}

scen_levels <- c("observaciones", "historical", "ssp126", "ssp245", "ssp370")
scen_labels <- c("Observado 1991-2020", "Histórico 1985-2014",
                 "SSP1-2.6 2041-70", "SSP2-4.5 2041-70", "SSP3-7.0 2041-70")
chill <- chill[scenario %in% scen_levels]
chill[, scenario := factor(scenario, levels = scen_levels, labels = scen_labels)]
```

```{r geo-base}
# Region de Murcia outline via mapSpain (downloaded and cached on first use). Fallback to a
# station bounding box if the download is unavailable, so the report still knits offline.
allst <- fread(STATIONS)
murcia <- tryCatch(st_transform(esp_get_prov("Murcia"), EPSG), error = function(e) NULL)
if (is.null(murcia)) {
  murcia <- st_as_sfc(st_bbox(st_transform(
    st_as_sf(allst, coords = c("lon", "lat"), crs = 4326), EPSG)))
}
muni <- tryCatch(st_transform(esp_get_munic(region = "Murcia"), EPSG), error = function(e) NULL)

# CORINE cropland backdrop: crop to Murcia, keep only strict (211-223) and heterogeneous
# (241-244) agriculture, and coarsen 100 m -> 1 km so it is a light context layer, not clutter.
clc <- rast(CLC)
clc_m <- mask(crop(clc, ext(vect(st_as_sf(murcia)))), vect(st_as_sf(murcia)))
rcl <- rbind(c(11.5, 17.5, 1), c(18.5, 22.5, 2))   # 12-17 -> 1 (puro), 19-22 -> 2 (heterogéneo)
crop_cat <- classify(clc_m, rcl, others = NA)
crop_cat <- aggregate(crop_cat, fact = 10, fun = "modal", na.rm = TRUE)
bd <- as.data.frame(crop_cat, xy = TRUE)
names(bd)[3] <- "cropland"
bd$cropland <- factor(round(bd$cropland), levels = c(1, 2),
                      labels = c("cultivo (211-223)", "heterogéneo (241-244)"))
bd <- bd[!is.na(bd$cropland), ]
```

## De qué va este informe

Test-run de la Región de Murcia: proyecciones AR6 de AEMET (ESD-RegBA, por estaciones)
convertidas a frío invernal con el modelo dinámico DM_JOSE (parametrización de 1987, script de
J.A. Egea; chill portions), y el Safe Winter Chill (percentil 10 de la acumulación de frío entre
inviernos) mapeado sobre el territorio. El caso
de estudio compara dónde se cumple el requerimiento de frío de 'Búlida' (47.5 CP) y del mutante
'Búlida Precoz' (33.7 CP). Corte de validación: 3 modelos que abarcan sensibilidad climática
alta, media y baja (UKESM1-0-LL, MPI-ESM1-2-HR, MIROC6), ventana de futuro cercano 2041-2070.

```{r fig1-context, fig.height=7}
# Fig 1: which stations sit on cultivable land, over the CORINE cropland backdrop.
strict <- fread(STRICT)[, .(station_id, strict = cultivable)]
broad  <- fread(BROAD)[, .(station_id, broad = cultivable)]
soil <- merge(merge(allst, strict, by = "station_id", all.x = TRUE), broad,
              by = "station_id", all.x = TRUE)
soil[, soil_cat := fifelse(strict %in% TRUE, "cultivable (estricto)",
                    fifelse(broad %in% TRUE, "solo amplio (heterogéneo)", "no cultivable"))]
soil[, soil_cat := factor(soil_cat,
     levels = c("cultivable (estricto)", "solo amplio (heterogéneo)", "no cultivable"))]
soil_sf <- st_transform(st_as_sf(soil, coords = c("lon", "lat"), crs = 4326), EPSG)

f1 <- ggplot() +
  geom_raster(data = bd, aes(x, y, fill = cropland)) +
  scale_fill_manual(values = c("cultivo (211-223)" = "#a1d99b",
                               "heterogéneo (241-244)" = "#fee08b"),
                    name = "Suelo agrícola\n(CORINE 2018)") +
  { if (!is.null(muni)) geom_sf(data = muni, fill = NA, color = "grey75", linewidth = 0.2) } +
  geom_sf(data = st_as_sf(murcia), fill = NA, color = "grey20", linewidth = 0.5) +
  geom_sf(data = soil_sf, aes(color = soil_cat), size = 2.4, stroke = 0.3) +
  scale_color_manual(values = c("cultivable (estricto)" = "#1a9850",
                                "solo amplio (heterogéneo)" = "#fd8d3c",
                                "no cultivable" = "#737373"), name = "Estación") +
  coord_sf(crs = EPSG, datum = NA) +
  labs(title = "Estaciones AEMET y suelo cultivable — Región de Murcia",
       subtitle = "criterio de referencia: ≥50% de cultivo en 2 km (44 estrictas, 73 amplias, de 123)") +
  theme_minimal(base_size = 12) +
  theme(axis.title = element_blank())
ggsave(file.path(FIG_DIR, "fig1_contexto_suelo.png"), f1, width = 9, height = 7.5, dpi = 150)
f1
```

## Safe Winter Chill por escenario (mediana de ensemble)

```{r fig2-swc}
# Ensemble collapse: median SWC across models per scenario and station ("ensemble, not mean").
ens <- chill[, .(swc_med = median(safe_winter_chill_P10),
                 n_models = .N,
                 n_ok_precoz = sum(safe_winter_chill_P10 >= CR_PRECOZ),
                 n_ok_bulida = sum(safe_winter_chill_P10 >= CR_BULIDA)),
             by = .(scenario, station_id, lon, lat)]
ens_sf <- st_transform(st_as_sf(ens, coords = c("lon", "lat"), crs = 4326), EPSG)

f2 <- ggplot() +
  geom_sf(data = st_as_sf(murcia), fill = "grey96", color = "grey55", linewidth = 0.25) +
  geom_sf(data = ens_sf, aes(color = swc_med), size = 1.7) +
  scale_color_viridis_c(name = "SWC\n(P10, CP)", option = "D") +
  facet_wrap(~scenario, nrow = 1) +
  coord_sf(crs = EPSG, datum = NA) +
  labs(title = "Safe Winter Chill: mediana de ensemble por escenario",
       subtitle = "chill portions acumuladas (P10 entre inviernos); 3 modelos, futuro 2041-2070") +
  theme_minimal(base_size = 11) +
  theme(axis.title = element_blank(), axis.text = element_blank())
ggsave(file.path(FIG_DIR, "fig2_swc_ensemble.png"), f2, width = 13, height = 5, dpi = 150)
f2
```

## Caso de estudio: ¿dónde cumple cada cultivar su requerimiento de frío?

```{r fig3-casestudy}
# The poster figure: classify each station (ensemble median SWC) against both cultivars' CR.
ens[, case := fifelse(swc_med >= CR_BULIDA, "cumplen ambas",
              fifelse(swc_med >= CR_PRECOZ, "solo 'Búlida Precoz'", "ninguna"))]
ens[, case := factor(case, levels = c("cumplen ambas", "solo 'Búlida Precoz'", "ninguna"))]
case_sf <- st_transform(st_as_sf(ens, coords = c("lon", "lat"), crs = 4326), EPSG)

f3 <- ggplot() +
  geom_sf(data = st_as_sf(murcia), fill = "grey96", color = "grey55", linewidth = 0.25) +
  geom_sf(data = case_sf, aes(color = case), size = 1.9) +
  scale_color_manual(values = c("cumplen ambas" = "#2166ac",
                                "solo 'Búlida Precoz'" = "#f4a582",
                                "ninguna" = "#b2182b"), name = "") +
  facet_wrap(~scenario, nrow = 1) +
  coord_sf(crs = EPSG, datum = NA) +
  labs(title = "¿Dónde se cumple el requerimiento de frío?",
       subtitle = "mediana de ensemble; 'Búlida' 47.5 CP, 'Búlida Precoz' 33.7 CP") +
  theme_minimal(base_size = 11) +
  theme(axis.title = element_blank(), axis.text = element_blank())
ggsave(file.path(FIG_DIR, "fig3_case_study.png"), f3, width = 13, height = 5, dpi = 150)
f3
```

```{r case-table}
# Share of stations in each class per scenario, as a quick read of the maps above.
tab <- ens[, .(estaciones = .N,
               cumplen_ambas = sum(case == "cumplen ambas"),
               solo_precoz   = sum(case == "solo 'Búlida Precoz'"),
               ninguna       = sum(case == "ninguna")), by = scenario]
knitr::kable(tab, caption = "Estaciones por clase y escenario (mediana de ensemble)")
```

## Distribución del Safe Winter Chill por escenario y modelo

```{r fig4-boxplots, fig.height=6}
# Model spread behind the ensemble, with each cultivar's CR as a reference line.
f4 <- ggplot(chill, aes(scenario, safe_winter_chill_P10, fill = model)) +
  geom_boxplot(outlier.size = 0.4, position = position_dodge(0.8), linewidth = 0.3) +
  geom_hline(yintercept = CR_BULIDA, color = "#2166ac", linewidth = 0.7) +
  geom_hline(yintercept = CR_PRECOZ, color = "#b2182b", linetype = "dashed", linewidth = 0.7) +
  annotate("text", x = 0.7, y = CR_BULIDA + 1.5, label = "Búlida 47.5", hjust = 0,
           size = 3, color = "#2166ac") +
  annotate("text", x = 0.7, y = CR_PRECOZ - 1.8, label = "B. Precoz 33.7", hjust = 0,
           size = 3, color = "#b2182b") +
  scale_fill_brewer(palette = "Set2", name = "Modelo") +
  labs(title = "Safe Winter Chill por escenario y modelo (todas las estaciones)",
       y = "SWC (P10, chill portions)", x = NULL) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
ggsave(file.path(FIG_DIR, "fig4_distribucion_swc.png"), f4, width = 10, height = 6, dpi = 150)
f4
```

```{r session, results='asis'}
cat("<hr><small>Figuras guardadas en <code>02_outputs/figures_chill/</code>. ",
    "Frío calculado con chillR (modelo dinámico DM_JOSE, param. 1987 de J.A. Egea) por <code>06_chill_murcia.R</code>; ",
    "datos PNACC AR6 ESD-RegBA extraídos con <code>02_pnacc_to_tables.py</code>.</small>", sep = "")
```
